Handling String-Number Conversion in String Constraint Solving

### 2019-11-14 draft.tex
don't know

The correctness of this construction is stated in Lemma~\ref{lemma:synccorrect} below. The lemma can also be understood as a precise specification the semantics of the synchronization formula, on which we rely for correctness of the construction of under-approximations of string constraints. 
%
\begin{lemma}\label{lemma:synccorrect}
%$\semof {\exists V_\times . \syncfof A {A'}}\restrict{\cvarone \cup \cvartwo \cup \pvarsof{\cvarone} \cup \pvarsof\cvartwo} = \{(I\cup\pim)\cup(I'\cup\pim') \mid \decode A(I\cup\pim) = \decode {A'}(I'\cup\pim')\}$. 
$\semof {\syncfof A {A'}}_{\cvarone\cup \cvartwo \cup \pvarone\cup\pvartwo} = \{(I\cup\pim)\cup(I'\cup\pim') \mid \decode A(I\cup\pim) = \decode {A'}(I'\cup\pim')\}$. 
\end{lemma}
%
Another version:
\begin{lemma}
$
\semof {\syncfof A {A'}}_{\cvarone\cup \cvartwo \cup \pvarone\cup\pvartwo} = 
\{
(I\cup\parikhof{x})\cup(I'\cup\parikhof{x'}) \mid 
x\in L(A), x'\in L(A'), I(x) = I(x')
\}
$
\end{lemma}
Notice that if $A$ is flat (or, symmetrically, if $A'$ is flat), then $\semof A \cap \semof {A'}$ can be still decoded from $\semof {\syncfof A {A'}}$ due to Corollary~\ref{corollary:pfa}, 
namely in the form of\\ $\semof A \cap \semof {A'} = \{\decode A(I\cup x) \mid \exists J.((I\cup x) \cup J) \in \semof{\syncfof A {A'}}\}$.  
\bigskip

@@

@@

\subsection{Handling Regular Constraints Symbolically} \label{section:mem}

\yfc{The version here is buggy, but fixable. I am not sure what is the best fix, so for now I just leave it here. The problem is we cannot synchronize a PFA and a general parametric automaton (which can be non-flat)}
Assume that the constraint we take from $\Phi_I$ is a regular constraint $\phi_r \defeq x \in L(A)$. Assume that $A=(Q,T,\Sigma,q_i,q_f)$. First we create a PFA $A' = (Q,T', V',q_i,q_f)$ that has exactly the same set of states as $A$. The set of transitions $T'$ and character variables $V'$ is defined as follows:

\begin{itemize}
	\item $V' = \{ v_{(q,a,q')} \mid (q,a,q')\in T\}$
	\item $T' = \{ (q, v_{(q,a,q'), q') } \mid (q,a,q')\in T\}$
\end{itemize}
We use the linear formula $\phi_c \defeq \bigwedge_{(q,a,q')\in T} {v_{(q,a,q')} = a}$ to remember the relation between the fresh character variables in $V'$ and the symbol it represents. The relation between $A$ and $A'$ is expressed in Lemma~\ref{lemma:fa_pfa}.

\begin{lemma}\label{lemma:fa_pfa}
	$L(A) = ( {\semof {\syncfof {A_x }{A'}} \cap \bigcap_{\phi\in\Phi_C} \semof {\phi} } )_{V_x \cup \#V_x} = \{(I\cup\pim) \mid \decode {A_x}(I\cup\pim) \in L(A)\}$. 
\end{lemma}

Intuitively, we create a new character variable for each transition in $T$, later we will add the relation $v_{(q,a,q')}  = a$ to $\Phi_O$.

The correctness of this construction is expressed in Lemma~\ref{lemma:memcorrect} below. 
%
%\newcommand\restrict[1]{\raisebox{-.5ex}{$\big|$}_{#1}}
\begin{lemma}\label{lemma:memcorrect}
	$({\semof {\syncfof {A_x }{A'}} \cap \bigcap_{\phi\in\Phi_C} \semof {\phi} } )_{V_x \cup \#V_x} = \{(I\cup\pim) \mid \decode {A_x}(I\cup\pim) \in L(A)\}$. 
\end{lemma}

@@

Assume that the constraint we take from $\Phi_I$ is a string-number conversion constraint  $\phi_s \defeq n=\sti{x}$. We will discuss how to make the corresponding linear formulas in this section.

Let us begin the discussion with a simple example, assume that we use the PFA in Figure~\ref{fig:sfa} (a) to restrict the domain of $x$. Then we know that when $0\leq v_0,v_1 \leq 9$, then $n$ is a positive integer value, and otherwise $n =-1$. So we should first add the constraint $ ((0\leq v_0\leq 9) \wedge (0\leq v_1\leq 9)) \vee n=-1 $.

@@

Our first attempt is a PFA with no loop, i.e., a straight line structure. The corresponding integer constraint does not have exponential components. However, it does not ensure the property that all numbers with at most $m$-digits are covered. Consider the example $\sti{x}=10 \wedge |x|=5$. The number $10$ has only $2$-digits, at the first glance, a straight-line PFA with two transitions, i.e., the PFA \scalebox{0.5}{\tikzset{state/.style={circle,draw=blue!50,fill=blue!20,
			thick,inner sep=0pt,minimum size=6mm}, initial text=$ $}
	\begin{tikzpicture} 
	\node[state,initial] (q0)  {$q_0$};
	\node[state] (q1) [right = 1cm of q0] {$q_1$};
	\node[state,accepting] (q2) [right = 1cm of q1] {$q_2$};

	\draw[->] (q0) edge node [below]{$v_1$} (q1) ;
	\draw[->] (q1) edge node [below]{$v_2$} (q2) ;
	\end{tikzpicture}} should be sufficient for the domain restriction of $x$. If we do so, we will conclude that the formula is unsatisfiable, because the length of $x$ cannot be $5$  under this domain restriction.

However, the formula is satisfiable when $x=``00010"$. Observe that $\sti{``00010"}=10$. The key is that even for a bounded integer, the corresponding string can be of unbounded length with arbitrarily many `$0$' characters at the front. In order to handle all number with at most $m$ digits, we propose to restrict the domain of $x$ to a PFA in Figure~\ref{fig:sfa_its}.  The PFA has a self-loop transition labeled $v_0$ at $q_0$ and we add a constraint $v_0=0$ to $\Phi_O$ for such PFA. With this transition, we can ensure that the under-approximation handles all numbers with at most $m$ digits. Although this PFA still has a loop, but it will not create any exponential component because the loop only represents a sequence of `0' that occurs in the prefix of $x$. It will not affect the integer value of $n=\sti{x}$.

@@

Such PFA handles bounded integers (e.g., 10-digits or 64-bits) precisely and under-approximate the possible values of unbounded integers. It works very efficiently because we just need to use linear constraints to characterize the relation between $x$ and $n$. 
Remember that the character variables can be assigned $\epsilon$. In order to simplify the construction, we an encoding to shift all the $\epsilon$-transitions to the least significant digits, i.e., if $v_j = \epsilon = -1$ then $v_k = -1$ for all $k\in [j+1,m]$.  To explain this encoding, we first define two macro function:

$$\begin{array}{rl}
\mathsf{epsilonAfter(k)}\defeq& (0\leq v_0,\ldots, v_k\leq 9) \wedge\\
&v_{k+1}=\ldots =v_m = -1\\
\mathsf{toNumber(k)}\defeq& (n = v_0\times 10^{k} + v_1\times 10^{k-1} + \ldots + v_{k})
\end{array}
$$

The function $\mathsf{epsilonAfter(k)}$ forces the first $k$ variables to be assigned some number and those after $k$ be assigned $\epsilon$. Recall that we use $-1$ to represent $\epsilon$. The function $\mathsf{toNumber(k)}$ converts the first $k$ variables to the corresponding number in $n$.
Then we can specify the relation between $n$ and the character variables as follows:
$$\phi_n\defeq (n=-1 \vee \bigvee_{k\in [0,m]} (\mathsf{epsAfter(k)} \wedge \mathsf{toNumber(k)} ) )$$
It says either $x$ is not a number and so $n=-1$ or the first $k$ variables represent the value of $n$, for some $k\in [0,m]$.

\yfc{add a Lemma for correctness}

@@

\section{Implementation}\label{section:implementation}
\input{evaluation}
\newpage
\bibliographystyle{ACM-Reference-Format}
\bibliography{refs}

>>>>

The correctness of this construction is stated in Lemma~\ref{lemma:synccorrect} below. The lemma can also be understood as a precise specification the semantics of the synchronization formula, on which we rely for correctness of the construction of under-approximations of string constraints. 
%
\begin{lemma}\label{lemma:synccorrect}
%$\semof {\exists V_\times . \syncfof A {A'}}\restrict{\cvarone \cup \cvartwo \cup \pvarsof{\cvarone} \cup \pvarsof\cvartwo} = \{(I\cup\pim)\cup(I'\cup\pim') \mid \decode A(I\cup\pim) = \decode {A'}(I'\cup\pim')\}$. 
$\semof {\syncfof A {A'}}_{\cvarone\cup \cvartwo \cup \pvarone\cup\pvartwo} = \{(I\cup\pim)\cup(I'\cup\pim') \mid \decode A(I\cup\pim) = \decode {A'}(I'\cup\pim')\}$. 
\end{lemma}
%
Another version:
\begin{lemma}
$
\semof {\syncfof A {A'}}_{\cvarone\cup \cvartwo \cup \pvarone\cup\pvartwo} = 
\{
(I\cup\parikhof{x})\cup(I'\cup\parikhof{x'}) \mid 
x\in L(A), x'\in L(A'), I(x) = I(x')
\}
$
\end{lemma}
Notice that if $A$ is flat (or, symmetrically, if $A'$ is flat), then $\semof A \cap \semof {A'}$ can be still decoded from $\semof {\syncfof A {A'}}$ due to Corollary~\ref{corollary:pfa}, 
namely in the form of\\ $\semof A \cap \semof {A'} = \{\decode A(I\cup \pim) \mid \exists J.((I\cup \pim) \cup J) \in \semof{\syncfof A {A'}}\}$.  
\bigskip

@@

\subsection{Handling Regular Constraints Symbolically} \label{section:mem}
Assume that the constraint we take from $\Phi_I$ is a regular constraint $\phi_r \defeq x \in L(A)$. Assume that $A=(Q,T,\Sigma,q_i,q_f)$. First we create a PFA $A' = (Q,T', V',q_i,q_f)$ that has exactly the same set of states as $A$. The set of transitions $T'$ and character variables $V'$ can be obtained by the following procedure.
Initially, both $V'$ and $T'$ are empty sets.
For each transition $(q, a, q') \in T$, we add a fresh character variable $v$ to $V'$, add the transition $(q,v,q')$ to $T'$, and add the constraint $v = \modelsof{a}$ to $\Phi_O$. Recall that $\modelsof{a}$ is the numerical encoding of the character $a$. After $A'$ is built, we add the synchronization formula $\syncfof {A_x }{A'}$ to $\Phi_O$.

The correctness of this construction is expressed in Lemma~\ref{lemma:memcorrect} below. 
%
%\newcommand\restrict[1]{\raisebox{-.5ex}{$\big|$}_{#1}}
\begin{lemma}\label{lemma:memcorrect}
	%$\semof {\exists V_\times . \syncfof A {A'}}\restrict{\cvarone \cup \cvartwo \cup \pvarsof{\cvarone} \cup \pvarsof\cvartwo} = \{(I\cup\pim)\cup(I'\cup\pim') \mid \decode A(I\cup\pim) = \decode {A'}(I'\cup\pim')\}$. 
	${\semof {\syncfof {A_x }{A'}}}_{V_x \cup \#V_x} = \{(I\cup\pim) \mid \decode {A_x}(I\cup\pim) \in L(A)\}$. 
\end{lemma}

@@

@@

In this section, we discuss how to handle the constraint $n=\sti{x}$ by restricting the domain of $x$ to some PFA. Let us begin with a simple example, assume that we use the PFA in Figure~\ref{fig:sfa} (a) to restrict the domain of $x$. Then we know that when $0\leq v_0,v_1 \leq 9$, then $n$ is a positive integer value, and otherwise $n =-1$. So we should first add the constraint $ ((0\leq v_0\leq 9) \wedge (0\leq v_1\leq 9)) \vee n=-1 $.

@@

Our first attempt is a PFA with no loop, i.e., a straight line structure (e.g., the PFA in Fig~\ref{fig:sfa_its} without the self-loop at $q_0$). The corresponding integer constraint does not have exponential components. However, it does not ensure the property that all numbers with at most $m$-digits are covered. Consider the example $\sti{x}=10 \wedge |x|=5$. The number $10$ has only $2$-digits, at the first glance, a straight-line PFA with two transitions, i.e., the PFA \scalebox{0.5}{\tikzset{state/.style={circle,draw=blue!50,fill=blue!20,
			thick,inner sep=0pt,minimum size=6mm}, initial text=$ $}
	\begin{tikzpicture} 
	\node[state,initial] (q0)  {$q_0$};
	\node[state] (q1) [right = 1cm of q0] {$q_1$};
	\node[state,accepting] (q2) [right = 1cm of q1] {$q_2$};

	\draw[->] (q0) edge node [below]{$v_1$} (q1) ;
	\draw[->] (q1) edge node [below]{$v_2$} (q2) ;
	\end{tikzpicture}}, should be sufficient for the domain restriction of $x$. If we do so, we will conclude that the formula is unsatisfiable, because the length of $x$ cannot be $5$  under this domain restriction.

However, the formula is satisfiable when $x=``00010"$. Observe that $\sti{``00010"}=10$. The key is that even for a bounded integer, the corresponding string can be of unbounded length with arbitrarily many `$0$' characters at the front. In order to handle all number with at most $m$ digits, we propose to restrict the domain of $x$ to a PFA in Figure~\ref{fig:sfa_its}.  The PFA has a self-loop transition labeled $v_0$ at $q_0$ and we always add a constraint $v_0=0$ for such PFA. With this transition, we can ensure that the under-approximation handles all numbers with at most $m$ digits. Although this PFA still has a loop, but it will not create any exponential component because the loop only represents a sequence of `0' that occurs in the prefix of $x$. It will not affect the integer value of $n=\sti{x}$.

@@

Such PFA handles bounded integers (e.g., 10-digits or 64-bits) precisely and under-approximate the possible values of unbounded integers. It works very efficiently because we just need to use linear constraints to characterize the relation between $x$ and $n$. 
Remember that the character variables can be assigned $\epsilon$. In order to handle such cases efficiently, we use an encoding that if $v_j = \epsilon = -1$ then $v_k = -1$ for all $k\in [j,m]$, i.e., we shift all the $\epsilon$-transitions to the least significant digits. In order to explain this encoding, we first define two macro function:

$$\begin{array}{rl}
\mathsf{epsilonAfter(k)}::=& (0\leq v_0,\ldots, v_k\leq 9) \wedge\\
&v_{k+1}=\ldots =v_m = -1\\
\mathsf{toNumber(k)}::=& (n = v_0\times 10^{k} + v_1\times 10^{k-1} + \ldots + v_{k})
\end{array}
$$

The function $\mathsf{epsilonAfter(k)}$ forces the first $k$ variables to be assigned some number and those after $k$ be assigned $\epsilon$. Recall that we use $-1$ to represent $\epsilon$. The function $\mathsf{toNumber(k)}$ converts the first $k$ variables to the corresponding number in $n$.
Then we can specify the relation between $n$ and the character variables as follows:
$$(n=-1 \vee \bigvee_{k\in [0,m]} (\mathsf{epsAfter(k)} \wedge \mathsf{toNumber(k)} ) )$$
It says either $x$ is not a number and so $n=-1$ or the first $k$ variables represent the value of $n$, for some $k\in [0,m]$ .

@@

\section{Implementation}\label{section:implementation}
\input{evaluation}

\bibliographystyle{ACM-Reference-Format}
\bibliography{refs}
